---
title: "The most ideal year"
author: "Markus Konrad"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
      number_sections: false
      toc: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtext)

theme_set(theme_minimal())
```


## TODOs

- write explanations
- proof-read
- update data
- remove this section
- deploy

## Introduction

Contemplating about the weather, I wondered if I could find out the "most unusual" and "most ideal" years regarding air temperature in Germany, i.e. if I could identify the years in which the daily temperature deviated the most and the least from the expected seasonal temperature. So I decided to look into historical climate data, created an extremely simplified seasonal temperature model and then investigated the deviations from that model. Although it's all quite simple, this little exploration gives some insights into how and why we can use a linear model for such data.

## Data

I retrieved the [historical climate data](https://www.dwd.de/DE/leistungen/klimadatendeutschland/klarchivtagmonat.html) for a weather station in Berlin-Dahlem (a bit outside the city) from 1950 to now from the [German Meteorological Service (Deutscher Wetterdienst – DWD)](https://www.dwd.de/). The data come as delimited files with semicolon as column separator. Historical data until 2022 and present data from 2022 to now come as separate files.

```{r}
raw_hist <- read.delim('data/produkt_klima_tag_19500101_20221231_00403.txt', sep = ';')
head(raw_hist)
```

```{r}
raw_pres <- read.delim('data/produkt_klima_tag_20230123_20240725_00403.txt', sep = ';')
head(raw_pres)
```

After reading in the files, we merge them, select only the necessary variables, transform the dates and remove duplicates (since the historical and the present data both contain observations from 2022) to generate our final measurements dataset `meas`:

```{r}
meas <- bind_rows(raw_hist, raw_pres) |>
    select(date = MESS_DATUM, temp = TMK) |>   # TMK is day-time average temperature in °C
    mutate(date = as.POSIXct(strptime(date, "%Y%m%d")),
           year = as.integer(as.numeric(format(date, "%Y"))),
           day = as.integer(as.numeric(format(date, "%j")))) |> # day of the year as decimal number from 1 to 366
    distinct(date, .keep_all = TRUE)   # remove duplicates
rm(raw_hist, raw_pres)                 # don't need the raw data any more
stopifnot(all(count(meas, date)$n == 1))   # make sure there are no duplicates
head(meas)
```

## Visual analysis

Let's visualize the time series with a simple plot. I will also add a smoothed curve showing an overall trend, which indicates a nearly linear increase in average yearly temperature by about 2°C since the 1950's. I'll later come back to that. We can also see the typical seasonal changes.

```{r temp-time-trend, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, fig.cap='Daily day-time average temperature in Berlin-Dahlem over time'}
ggplot(meas, aes(date, temp)) +
    geom_line() +
    geom_smooth(method = "gam") +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
         x = "",
         y = "Temperature in °C")
```

The periodical temperature changes can be visualized by by looking at a smaller time frame:

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
filter(meas, year >= 2018) |>
    ggplot(aes(date, temp)) +
        geom_line() +
        geom_smooth(span = 0.2, method = "loess") +
        labs(title = "Daily day-time average temperature in Berlin-Dahlem since 2018",
             x = "",
             y = "Temperature in °C")
```
<!--
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
filter(meas, year == 2023) |>
    ggplot(aes(date, temp)) +
        geom_line(method = "loess") +
        geom_smooth() +
        labs(title = "Daily day-time average temperature in Berlin-Dahlem in 2023",
             x = "",
             y = "Temperature in °C")
```
-->

We can also plot the yearly trend by plotting the temperature against the day of the year. We can see the typical seasonal pattern but also the slight overall increase in temperature over the years, since more recent years (yellow color) tend to have higher temperatures, especially in the winter.

```{r temp-time-seasonal, fig.width=10, fig.height=6, fig.cap='Daily day-time average temperature in Berlin-Dahlem over time'}
ggplot(meas, aes(day, temp, color = year)) +
    geom_line(alpha = 0.25) +
    scale_color_binned(name = "Year", type = 'viridis') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
             x = "Day of the year",
             y = "Temperature in °C")
```

## Modeling

Naturally, and confirmed with the above plots, we can use a periodic function like the cosine function to model these temperatures. In general this periodic function can be written as

\begin{equation}
y = c \cos (x + \varphi), (\#eq:periodic)
\end{equation}

where $c$ controls the amplitude (maximum spikes), $\varphi$ the phase (shift on the x-axis) and $x$ the frequency. With a linear model, we can only fit linear terms like $y = ax + b$, so we have the problem that we can't estimate the frequency and the phase. Luckily – in our very simple case – the frequency is already known: the seasonal pattern repeats yearly, so we can calculate $x = 2 \pi D / 366$, where $D$ is the day of the year. Because of leap years, $D$ can range from 1 to 366 and so we divide it by 366. This means that over the course of a year, $x$ makes "a full circle" from above 0 to $2 \pi$.

The second problem – that we can't estimate the phase with a linear model directly – can be solved by applying a neat trick that [transforms the cosine wave with an amplitude and a phase shift to a linear combination of a cosine and a sine wave](https://en.wikipedia.org/wiki/List_of_trigonometric_identities#Linear_combinations):

\begin{equation}
c \cos (x + \varphi) = a \cos x + b \sin x, (\#eq:linearcomb)
\end{equation}

where

$$
c = \text{sgn}(a) \sqrt{a^2 + b^2}, \\
\varphi = \arctan \frac{-b} a.
$$

This means we can estimate $a$ and $b$ as coefficients for the above linear combination that is equivalent to the initially defined cosine wave. Hence we can finally specify our linear model `m1` for the temperatures $Y_t$ as


\begin{equation}
Y_t = \beta_0 + \beta_1 \cos(x_t) + \beta_2 \sin(x_t) + \epsilon_t,  (\#eq:m1)
\end{equation}

where $x_t$ is the only regressor – the day of the year transformed to range $(0, 2 \pi]$ as described above –, $\beta_0$ to $\beta_2$ are the coefficients we seek to estimate and  $\epsilon_t$ is the error term.

## Model estimation

We can now estimate the model `m1` in R by first computing $x_t$ and then using `lm` to fit the model using our measurements `meas`:

```{r}
# compute frequency x
meas <- mutate(meas, x = 2 * pi * day/366)

# fit the model
m1 <- lm(temp ~ cos(x) + sin(x), meas)
summary(m1)
```

The model has a decent fit given its simplicity: About 75% of the variation of the temperature can be explained by the seasonal pattern we modeled with the periodic functions. The estimated intercept reflects the mean of the temperature and the coefficients for the periodic functions result in the oscillation around the mean.

We can see that for predicted values near 0°C the errors are a bit larger and also the distribution of the errors is slightly left skewed.

```{r}
plot(m1, which = 1:2, ask = FALSE)
```

Let's plot the model predictions of the temperatures on top of the measured values. For a better overview, we only consider data from 2018 or newer:

```{r, fig.width=10, fig.height=6}
meas_fit <- cbind(meas, pred = fitted(m1))

filter(meas_fit, year >= 2018) |>
    ggplot() +
        geom_line(aes(date, temp), alpha = 0.25) +
        geom_line(aes(date, pred), color = 'red') +
        labs(title = "Daily day-time average temperature in Berlin-Dahlem since 2018",
             subtitle = "Model predictions in <span style='color:red'>red</span>.",
             x = "",
             y = "Temperature in °C") +
        theme(plot.subtitle = element_markdown())
```

Let's show that we can again apply the transformation in equation \@ref(eq:linearcomb) so that we can recover the cosine wave form in equation \@ref(eq:periodic) from the linear combination form of the model equation \@ref(eq:m1). As expected, we get same predictions from both model representations:

```{r, fig.width=10, fig.height=6}
a <- m1$coefficients[2]
b <- m1$coefficients[3]

# calculate c and phi from the coefficients a and b
c <- sign(a) * sqrt(a^2 + b^2)
phi <- atan(-b/a)

# predictions from the cosine wave form of the model equation for model m1
meas_fit$pred2 <- m1$coefficients[1] + c * cos(meas$x + phi)

filter(meas_fit, year >= 2018) |>
    ggplot() +
        geom_line(aes(date, temp), alpha = 0.25) +
        geom_line(aes(date, pred), color = 'red') +
        geom_line(aes(date, pred2), color = 'blue', linetype = "dashed") +
        labs(title = "Daily day-time average temperature in Berlin-Dahlem since 2018",
             subtitle = "Model predictions in <span style='color:red'>red (linear combination form)</span> 
                         and <span style='color:blue'>blue (cosine wave form)</span>.",
             x = "",
             y = "Temperature in °C") +
        theme(plot.subtitle = element_markdown())

```

Let's plot the model predictions for the whole time range.

```{r, fig.width=10, fig.height=6}
ggplot(meas_fit) +
    geom_line(aes(date, temp), alpha = 0.25) +
    geom_line(aes(date, pred), color = 'red') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
         subtitle = "Model predictions in <span style='color:red'>red</span>.",
         x = "",
         y = "Temperature in °C") +
    theme(plot.subtitle = element_markdown())
```

It's barely visible at this scale, but the model systematically overestimates temperatures around the beginning of the time range and underestimates temperatures towards the end of the time range. We can see this more clearly with a plot similar to figure \@ref(fig:temp-time-seasonal). Temperatures before 1970 (dark purple) tend to be overestimated while temperatures from 2000 and later tend to be underestimated:

```{r, fig.width=10, fig.height=6}
ggplot(meas_fit, aes(day, temp, color = year)) +
    geom_line(alpha = 0.25) +
    geom_line(aes(day, pred), color = 'red') +
    scale_color_binned(name = "Year", type = 'viridis') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
         subtitle = "Model predictions in <span style='color:red'>red</span>.",
         x = "Day of the year",
         y = "Temperature in °C") +
    theme(plot.subtitle = element_markdown())
```

We can confirm this conjecture quantitatively by calculating the mean error of our predictions $\hat Y_t - Y_t$ for different decades:

```{r mean-pred-err-decade-m1, fig.cap="Mean prediction error per decade for m1"}
(mean_err_per_decade <- mutate(meas_fit, d = round((year - 1900) / 10) * 10,
                 decade = as.ordered(ifelse(d < 100, paste0("19", d, "s"), sprintf("20%02ds", d - 100)))) |>
    group_by(decade) |>
    summarise(mean_error = mean(pred - temp)))
ggplot(mean_err_per_decade, aes(x = decade, y = mean_error)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point() +
    scale_y_continuous(limits = c(-2, 2)) +
    labs(title = "Mean prediction error per decade for m1",
         x = "Decade",
         y = "Prediction error in °C")
```

## Improving the model

The reason for these systematic errors is the slight increase of the yearly average temperature as indicated by the trend line in figure \@ref(fig:temp-time-trend). To put it bluntly: we didn't account for global warming! We can do so by updating the model equation from model `m1` \@ref(eq:m1) in order to include a term for a yearly linear change (i.e. increase) in temperature. Of course this is very simplified, but it should improve our previous model. So our updated model `m2` is now:

$$
Y_t = \beta_0 + \beta_1 \cos(x_t) + \beta_2 \sin(x_t) + \beta_3 \text{year}_t + \epsilon_t.
$$

And we can fit this updated model:

```{r}
m2 <- lm(temp ~ cos(x) + sin(x) + year, meas)
summary(m2)
```

We can see that the fit is slightly better. $R^2$ increased by about $1$ percentage points, indicating that about $1\%$ of the variation of the average daily temperature can be attributed to the yearly temperature increase. This doesn't sound much, however, you should keep in mind that we're talking about daily average temperatures which are influenced by a lot of factors.

The estimated coefficient for the yearly change is about $0.0289$, which means that the we expect the yearly average temperature to increase by about $0.289\text{°C}$ within $10$ years, or about $2.89\text{°C}$ within $100$ years, which is indeed quite much.

Note also that our intercept has changed from about $9.4\text{°C}$ to about $-47.3\text{°C}$, because the model needs to account for the yearly temperature increase and hence predicts an average temperature of $-47.3\text{°C}$ in year $0$, while for example the average temperature in year $2000$ is predicted as $-47.3\text{°C} + 2000 \cdot 0.0289\text{°C} = 10.5\text{°C}$ (and $39.4\text{°C}$ in year $3000$!). This of course shows that such simple models should never be used for extrapolation.

The model fit inspection plots didn't change much:

```{r}
plot(m2, which = 1:2, ask = FALSE)
```

We again check the model fit visually for a smaller time span:

```{r, fig.width=10, fig.height=6}
meas_fit2 <- cbind(meas, pred = fitted(m2))

filter(meas_fit2, year >= 2018) |>
    ggplot() +
        geom_line(aes(date, temp), alpha = 0.25) +
        geom_line(aes(date, pred), color = 'red') +
        labs(title = "Daily day-time average temperature in Berlin-Dahlem since 2018",
             subtitle = "Model predictions in <span style='color:red'>red</span>.",
             x = "",
             y = "Temperature in °C") +
        theme(plot.subtitle = element_markdown())
```

And we also check it for the full time range. Here, we also see the linear increase in average temperature introduced in the updated model:

```{r, fig.width=10, fig.height=6}
ggplot(meas_fit2) +
    geom_line(aes(date, temp), alpha = 0.25) +
    geom_line(aes(date, pred), color = 'red') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
         subtitle = "Model predictions in <span style='color:red'>red</span>.",
         x = "",
         y = "Temperature in °C") +
    theme(plot.subtitle = element_markdown())
```

```{r, fig.width=10, fig.height=6}
ggplot(meas_fit2, aes(day, temp, color = year)) +
    geom_line(alpha = 0.25) +
    geom_line(aes(day, pred, color = year, group = year)) +
    scale_color_binned(type = 'viridis') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time",
         subtitle = "Model predictions as opaque colored curves.",
         x = "Day of the year",
         y = "Temperature in °C") +
    theme(plot.subtitle = element_markdown())
```

As compared to model `m1` (see figure \@ref(fig:mean-pred-err-decade-m1)), we don't see the systematic underestimation of the temperature in the later decades anymore:

```{r mean-pred-err-decade-m2, fig.cap="Mean prediction error per decade for m2"}
(mean_err_per_decade2 <- mutate(meas_fit2, d = round((year - 1900) / 10) * 10,
                                           decade = as.ordered(ifelse(d < 100, paste0("19", d, "s"),
                                                                      sprintf("20%02ds", d - 100)))) |>
    group_by(decade) |>
    summarise(mean_error = mean(pred - temp)))

ggplot(mean_err_per_decade2, aes(x = decade, y = mean_error)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point() +
    scale_y_continuous(limits = c(-2, 2)) +
    labs(title = "Mean prediction error per decade for m2",
         x = "Decade",
         y = "Prediction error in °C")
```

## Investigating deviations from the model

The `m2` model seems good enough for our purpose to identify the years with the least and the most deviation from the typical seasonal temperatures. We will hence analyze the residuals $Y_t - \hat Y_t$ from `m2` and we'll start by looking at their distribution which is, as confirmed earlier by the model inspection plots, normal around 0°C:

```{r, fig.width=6, fig.height=4}
resid <- meas_fit2$temp - meas_fit2$pred
ggplot(data.frame(resid = resid), aes(resid, after_stat(density))) +
    geom_histogram(bins = 20) +
    labs(title = "Distribution of residuals for model m2",
         x = "Residual of predicted temperature in °C",
         y = "Density")
```

We want to investigate the yearly rate of "unusual" day temperatures, i.e. on how many days the temperature was either much lower or much higher than the typical temperature for that season predicted by our model `m2`. Choosing a threshold for this can be quite arbitrary. We will at least base the threshold on the given historical data, by using the 90th-percentile of the absolute residuals as threshold:

```{r}
quantile(abs(resid), 0.9)
```

This means an "unusual" temperature is one that, from our historical data, is either higher or lower than the seasonal trend plus the global warming trend by around $6\text{°C}$ at 10% of the days  since 1950. We now use this threshold to calculate some statistics from the residuals for each year: The mean deviation $\overline{Y_t - \hat Y_t}$, mean absolute deviation $\overline{|Y_t - \hat Y_t|}$ and the proportions of unusually cold or warm days.

```{r, fig.width=10, fig.height=6}
thresh_unusal_temp <- 6

resid_stats <- group_by(meas_fit2, year) |>
    summarise(me = mean(temp - pred),
              mae = mean(abs(temp - pred)),
              prop_days_warmer = mean(temp > pred + thresh_unusal_temp),
              prop_days_colder = mean(temp < pred - thresh_unusal_temp))

resid_stats_plt <- pivot_longer(resid_stats, !year, names_to = "measure")

filter(resid_stats_plt, measure %in% c("mae", "me")) |>
    mutate(measure = factor(case_match(measure, "mae" ~ "Mean abs. deviation", "me" ~ "Mean deviation"),
                            levels = c("Mean deviation", "Mean abs. deviation"))) |>
    ggplot(aes(x = year, y = value, fill = measure)) +
        geom_col(position = position_dodge()) +
        scale_fill_discrete(guide = "none") +
        facet_wrap(vars(measure), nrow = 2, scales = "free_y") +
        labs(title = "Deviation from the typical seasonal temperature",
             x = "Year",
             y = "Temperature in °C")
```

We can see that there's quite some variation in terms of mean deviation in the different years. Some years like 1996 or 2010 tended to be mostly colder than usual, while 1953 and 2024 (so far) were mostly warmer than usual:

```{r}
select(resid_stats, year, me) |>
    arrange(me) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, everything())
```

Exceptionally high and low temperatures may cancel each other out, so we have to look at the mean absolute deviation and at the proportion of unusual cold or warm days. 2017, 1973 and 1958 are the years with lowest mean absolute deviation from the seasonal trend, while 1956, 2010 and 1985 deviated the most.

```{r}
select(resid_stats, year, mae) |>
    arrange(mae) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, everything())
```


```{r, fig.width=10, fig.height=6}
filter(resid_stats_plt, measure %in% c("prop_days_warmer", "prop_days_colder")) |>
    mutate(measure = case_match(measure, "prop_days_warmer" ~ "Warmer", "prop_days_colder" ~ "Colder")) |>
    ggplot(aes(x = year, y = value, fill = measure)) +
        geom_col(position = position_stack()) +
        scale_fill_discrete(limits = rev, name = "") +
        labs(title = "Proportion of unusually warm or cold days",
             x = "Year",
             y = "")
```

1958, 2017 and 2023 had the least number of days with unusually low or high temperatures and 1985, 2010 and 2013 had the most. Again, please keep in mind, that our model `m2` already takes into account the yearly temperature increase caused by global warming, so "unusual" refers only to the seasonal trend.

```{r}
select(resid_stats, year, prop_days_warmer, prop_days_colder) |>
    mutate(prop_days_unusual = prop_days_warmer + prop_days_colder) |>
    arrange(prop_days_unusual) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, year, prop_days_unusual, everything())
```

We identify the years with the least and most deviation using the the mean absolute deviation measure. By this measure, 2017 is the "most ideal year", i.e. the year following closest the seasonal trend. This is also reflected by a small proportion of unusual day temperatures (ranking first in a tie with year 1958 in the above table). In contrast, 1985 is the year with the highest mean absolute deviation. It also has the highest proportion of unusual day temperatures, especially caused by a very cold winter. The following figure compares both years and highlights unusual temperatures:

```{r, fig.width=10, fig.height=6}
least_deviation_yr <- slice_min(resid_stats, mae) |> pull(year)
most_deviation_yr <- slice_max(resid_stats, mae) |> pull(year)

least_most_plt <- data.frame(year = c(least_deviation_yr, most_deviation_yr), label = c("least deviation", "most deviation")) |>
    inner_join(meas_fit2, by = 'year') |>
    mutate(label = paste0(year, " (", label, ")"),
           resid = temp - pred,
           transparency = ifelse(abs(resid) > thresh_unusal_temp, 0.5, 0.1))

ggplot(least_most_plt, aes(day, temp, color = label)) +
    geom_point(aes(alpha = transparency)) +
    geom_line(aes(day, pred)) +
    scale_color_discrete(guide = guide_legend(title = NULL)) +
    scale_alpha_identity(guide = NULL) +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time for select years",
         x = "Day of the year",
         y = "Temperature in °C")
```

Looking only at the residuals gives a clearer image regarding the deviations:

```{r, fig.width=10, fig.height=4}
ggplot(least_most_plt, aes(day, resid, color = label, alpha = transparency)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_hline(yintercept = -thresh_unusal_temp, linetype = "dotted") +
    geom_hline(yintercept = thresh_unusal_temp, linetype = "dotted") +
    geom_point() +
    scale_color_discrete(guide = guide_legend(title = NULL)) +
    scale_alpha_identity(guide = NULL) +
    labs(title = "Residuals for select years",
         x = "Day of the year",
         y = "Residual of predicted temperature in °C")
```

## Global warming

```{r}
# fit the model
m3 <- lm(temp ~ cos(x) + sin(x), filter(meas, year <= 1959))
summary(m3)
```

```{r, fig.width=10, fig.height=6}
meas_fit3 <- cbind(meas, pred = predict(m3, meas))

ggplot(meas_fit3) +
    geom_line(aes(date, temp), alpha = 0.25) +
    geom_line(aes(date, pred), color = 'red') +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem",
         subtitle = "Model predictions in <span style='color:red'>red</span>.",
         x = "",
         y = "Temperature in °C") +
    theme(plot.subtitle = element_markdown())
```

```{r}
(mean_err_per_decade3 <- mutate(meas_fit3, d = round((year - 1900) / 10) * 10,
                                           decade = as.ordered(ifelse(d < 100, paste0("19", d, "s"),
                                                                      sprintf("20%02ds", d - 100)))) |>
    group_by(decade) |>
    summarise(mean_error = mean(pred - temp)))

ggplot(mean_err_per_decade3, aes(x = decade, y = mean_error)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point() +
    scale_y_continuous(limits = c(-2, 2)) +
    labs(title = "Mean prediction error per decade for m3",
         x = "Decade",
         y = "Prediction error in °C")
```

```{r, fig.width=10, fig.height=6}
resid_stats <- group_by(meas_fit3, year) |>
    summarise(me = mean(temp - pred),
              mae = mean(abs(temp - pred)),
              prop_days_warmer = mean(temp > pred + thresh_unusal_temp),
              prop_days_colder = mean(temp < pred - thresh_unusal_temp))

resid_stats_plt <- pivot_longer(resid_stats, !year, names_to = "measure")

filter(resid_stats_plt, measure %in% c("mae", "me")) |>
    mutate(measure = factor(case_match(measure, "mae" ~ "Mean abs. deviation", "me" ~ "Mean deviation"),
                            levels = c("Mean deviation", "Mean abs. deviation"))) |>
    ggplot(aes(x = year, y = value, fill = measure)) +
        geom_col(position = position_dodge()) +
        scale_fill_discrete(guide = "none") +
        facet_wrap(vars(measure), nrow = 2, scales = "free_y") +
        labs(title = "Deviation from the typical seasonal temperature",
             x = "Year",
             y = "Temperature in °C")
```

```{r}
select(resid_stats, year, me) |>
    arrange(me) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, everything())
```

```{r}
select(resid_stats, year, mae) |>
    arrange(mae) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, everything())
```


```{r, fig.width=10, fig.height=6}
filter(resid_stats_plt, measure %in% c("prop_days_warmer", "prop_days_colder")) |>
    mutate(measure = case_match(measure, "prop_days_warmer" ~ "Warmer", "prop_days_colder" ~ "Colder")) |>
    ggplot(aes(x = year, y = value, fill = measure)) +
        geom_col(position = position_stack()) +
        scale_fill_discrete(limits = rev, name = "") +
        labs(title = "Proportion of unusually warm or cold days",
             x = "Year",
             y = "")
```

```{r}
select(resid_stats, year, prop_days_warmer, prop_days_colder) |>
    mutate(prop_days_unusual = prop_days_warmer + prop_days_colder) |>
    arrange(prop_days_unusual) |>
    mutate(rank = row_number()) |>
    filter(rank %in% c(1:3, (n()-2):n())) |>
    select(rank, year, prop_days_unusual, everything())
```


```{r}
select(resid_stats, year, prop_days_warmer) |>
    arrange(desc(prop_days_warmer)) |>
    head(5)
```


```{r, fig.width=10, fig.height=6}
least_deviation_yr <- slice_min(resid_stats, mae) |> pull(year)
most_deviation_yr <- slice_max(resid_stats, mae) |> pull(year)

least_most_plt <- data.frame(year = c(least_deviation_yr, most_deviation_yr), label = c("least deviation", "most deviation")) |>
    inner_join(meas_fit3, by = 'year') |>
    mutate(label = paste0(year, " (", label, ")"),
           resid = temp - pred,
           transparency = ifelse(abs(resid) > thresh_unusal_temp, 0.5, 0.1))

ggplot(least_most_plt) +
    geom_point(aes(day, temp, alpha = transparency, color = label)) +
    geom_line(aes(day, pred)) +
    scale_color_discrete(guide = guide_legend(title = NULL)) +
    scale_alpha_identity(guide = NULL) +
    labs(title = "Daily day-time average temperature in Berlin-Dahlem over time for select years",
         x = "Day of the year",
         y = "Temperature in °C")
```

## Conclusion

TODO
